---
title: "Generate synthetic EHR data"
author: "Jun Ha Chang"
created: "2026-06-29"
last_updated: "2026-06-30"
categories: [R, EHR]
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(lubridate)
library(stringr)

set.seed(2026)

current_dir <- getwd()
```

*Created: 2026-06-29* *Last update: 2026-06-30*

# PURPOSE

This script generates synthetic EHR-style data for practicing cohort feasibility logic. The data are fully synthetic and are intended to mimic common relational EHR structures.

# PATIENT

The patient table contains one row per individual and stores demographic information and death date. This table will later be used to calculate age, identify missing baseline covariates, and define censoring due to death.

-   PATID
-   BIRTH_DATE
-   DEATH_DATE
-   SEX
-   RACE

## Parameter setup

These parameters control the number of synthetic patients, demographic distributions, observation window, and death rate. Changing these values will change the size and structure of the generated synthetic cohort.

```{r}
patient_n <- 100 # 100 patients 
sex_ratio <- 0.5 # 50% female, 50% male 

race_levels <- c("White", "Black", "Asian", "Other")
race_dist <- c(0.60, 0.14, 0.07, 0.19)

age_mean <- 65 # mean age of patients
age_sd <- 15 # standard deviation of age

data_start <- as.Date("2016-01-01") # Date of starting EHR data 
data_end   <- as.Date("2026-06-29") # Date of query data

death_rate <- 0.05   # 5% have recorded death date
```

## Generate synthenic data

This chunk creates patient IDs, birth dates, sex, race, and death dates. The final output is a patient-level table with one row per PATID.

```{r}
patients <- tibble(
  PATID = str_c("P", str_pad(1:patient_n, width = 4, pad = "0")),
  
  AGE_AT_DATA_END = round(rnorm(patient_n, mean = age_mean, sd = age_sd)), 
  
  BIRTH_DATE = data_end - years(AGE_AT_DATA_END) - days(sample(0:364, patient_n, replace = TRUE)), 
  
  SEX = sample(
    c("Female", "Male"),
    size = patient_n,
    replace = TRUE, 
    prob = c(sex_ratio, 1 - sex_ratio)
  ),
  
  RACE = sample(
    race_levels, 
    size = patient_n,
    replace = TRUE, 
    prob = race_dist
  ),
  
  DEATH = rbinom(patient_n, size = 1, prob = death_rate), 
  
  DEATH_DATE = if_else(
    DEATH == 1, 
    sample(seq.Date(data_start + years(1), data_end, by = "day"),
           size = patient_n, replace = TRUE), 
    as.Date(NA)
  )
) %>%
  select(PATID, BIRTH_DATE, DEATH_DATE, SEX, RACE)

write.csv(patients, "synthetic_patients.csv", row.names = FALSE, na = "")

```

# DIAGNOSES

The diagnosis table contains multiple rows per patient. This reflects typical EHR data, where a patient may have repeated diagnosis records over time.

-   PATID
-   ICD10
-   ICD10_TXT
-   DX_DATE

## Parameter setup

This chunk defines the ICD-10 code dictionary and diagnosis-generation parameters. Sleep disorder diagnoses will be used to define the index date, and Alzheimer's disease diagnoses will be used to define prevalent versus incident outcomes.

```{r}
# Diagnosis code dictionary
dx_dictionary <- tibble::tribble(
  ~ICD10,   ~ICD10_TXT,   
  # Sleep disorders
  "G47.00", "Insomnia disorder",                        
  "G47.33", "Obstructive sleep apnea",                  
  "G47.9",  "Sleep disorder, unspecified",              
  
  # Alzheimer's disease
  "G30.0",  "Alzheimer's disease with early onset",     
  "G30.1",  "Alzheimer's disease with late onset",      
  "G30.9",  "Alzheimer's disease, unspecified",         
  
  # Common comorbidities for older adults population
  "I10",    "Essential hypertension",                      
  "E78.5",  "Hyperlipidemia, unspecified",                  
  "M19.90", "Osteoarthritis, unspecified site",             
  "E11.9",  "Type 2 diabetes mellitus without complications",
  "I25.10", "Atherosclerotic heart disease of native coronary artery without angina pectoris"
                        
)

# Parameter setup
sleep_dx_rate <- 0.55       # 55% have at least one sleep disorder diagnosis
ad_dx_rate <- 0.18          # 18% have Alzheimer's disease diagnosis
prevalent_ad_rate <- 0.25   # among AD cases, 25% occur before sleep diagnosis
mean_comorbid_dx <- 2       # average number of unrelated diagnoses per patient

comorbidity_codes <- c("I10", "E78.5", "M19.90", "E11.9", "I25.10")
comorbidity_prob <- c(0.35, 0.25, 0.18, 0.15, 0.07)
```

## Generate synthenic data

### Generate sleep disorder diagnoses

This chunk randomly selects patients with sleep disorder diagnoses and assigns each selected patient one sleep-related ICD-10 code and diagnosis date.

```{r}
# Select individuals with sleep disorder diagnoses
sleep_pat <- patients %>%
  slice_sample(prop = sleep_dx_rate) %>% # Randomly select individuals
  pull(PATID)

sleep_dx <- tibble(
  PATID = sleep_pat,
  ICD10 = sample(
    c("G47.00", "G47.33", "G47.9"),
    size = length(sleep_pat),
    replace = TRUE,
    prob = c(0.40, 0.45, 0.15) # Modify probabilities for specific sleep disorder dx
  ),
  DX_DATE = sample(
    seq.Date(data_start, data_end, by = "day"),
    size = length(sleep_pat),
    replace = TRUE
  )
) %>%
  left_join(dx_dictionary, by = "ICD10") %>%
  select(PATID, ICD10, ICD10_TXT, DX_DATE)
```

### Generate Alzheimer's disease diagnoses

This chunk randomly selects patients with Alzheimer's disease diagnoses. In the later feasibility workflow, Alzheimer's disease dates will be compared with sleep disorder index dates to distinguish prevalent from incident Alzheimer's disease.

```{r}
# Select individuals with Alzheimer's disease 
ad_pat <- patients %>%
  slice_sample(prop = ad_dx_rate) %>% # Randomly select individuals
  pull(PATID)

ad_dx <- tibble(PATID = ad_pat,
                 ICD10 = sample(
                   c("G30.0", "G30.1", "G30.9"),
                   size = length(ad_pat),
                   replace = TRUE,
                   prob = c(0.20, 0.50, 0.30) # Modify probabilities for specific AD dx
                 ),
                 DX_DATE = sample(
                   seq.Date(data_start, data_end, by = "day"),
                   size = length(ad_pat),
                   replace = TRUE
                 )
) %>%
  left_join(dx_dictionary, by = "ICD10") %>%
  select(PATID, ICD10, ICD10_TXT, DX_DATE)
```

### Generate comorbidity diagnoses

This chunk adds common chronic conditions for older adults. These records make the diagnosis table look more like real EHR data, where patients often have several comorbid diagnoses.

```{r}
# Generate common comorbidity diagnosis records
other_dx <- patients %>%
  select(PATID) %>%
  rowwise() %>%
  mutate(n_comorbid_dx = rpois(1, lambda = mean_comorbid_dx)) %>%
  ungroup() %>%
  filter(n_comorbid_dx > 0) %>%
  tidyr::uncount(n_comorbid_dx) %>%
  mutate(
    ICD10 = sample(
      comorbidity_codes,
      size = n(),
      replace = TRUE,
      prob = comorbidity_prob
    ),
    DX_DATE = sample(
      seq.Date(data_start, data_end, by = "day"),
      size = n(),
      replace = TRUE
    )
  ) %>%
  left_join(dx_dictionary, by = "ICD10") %>%
  select(PATID, ICD10, ICD10_TXT, DX_DATE)
```

### Combine diagnosis records

This chunk stacks sleep disorder, Alzheimer's disease, and comorbidity records into one diagnosis table. The final `dx` table is intentionally many-to-one: each patient can contribute multiple diagnosis records.

```{r}
dx <- bind_rows(
  sleep_dx,
  ad_dx,
  other_dx
) %>%
  arrange(PATID, DX_DATE, ICD10)

# Preview
dx %>%
  slice_head(n = 15)

# Output 
write.csv(dx, "synthetic_diagnoses.csv", row.names = FALSE, na = "")
```

# ENCOUNTER

The encounter table contains multiple rows per patient and represents observable EHR activity. It will later be used to calculate baseline lookback and follow-up time.

-   PATID
-   ENCOUNTER_ID
-   ENCOUNTER_DATE
-   ENCOUNTER_TYPE

## Parameter setup

These parameters control the average number of encounters per patient and the distribution of encounter types.

```{r}
mean_encounters_per_patient <- 8

encounter_type_levels <- c(
  "Outpatient",
  "Primary care",
  "Specialty care",
  "ED",
  "Inpatient"
)

encounter_type_prob <- c(0.45, 0.25, 0.15, 0.08, 0.07)

```

## Generate synthenic data

### Generate synthetic encounter data

This chunk expands each patient into multiple encounter records, assigns encounter dates and encounter types, and removes encounters occurring after death. The final table is encounter-level, not patient-level.

```{r}
encounter <- patients %>%
  select(PATID, DEATH_DATE) %>%
  rowwise() %>%
  mutate(
    encounter_n = max(1, rpois(1, lambda = mean_encounters_per_patient))
    ) %>% # Ensure at least 1 encounter
  ungroup() %>%
  tidyr::uncount(encounter_n) %>%
  group_by(PATID) %>%
  mutate(
    ENCOUNTER_DATE = sample(
      seq.Date(data_start, data_end, by = "day"),
      size = n(), 
      replace = TRUE
    ), 
    ENCOUNTER_TYPE = sample(
      encounter_type_levels,
      size = n(),
      replace = TRUE,
      prob = encounter_type_prob
  )) %>%
  ungroup() %>%
  # Remove encounnters that occur after death date
  filter(is.na(DEATH_DATE) | ENCOUNTER_DATE <= DEATH_DATE) %>%
  arrange(PATID, ENCOUNTER_DATE) %>%
  mutate(
    ENCOUNTER_ID = str_c("E", str_pad(row_number(), width = 3, pad = "0"))
  ) %>% 
  select(-DEATH_DATE)
  
# Preview
encounter %>%
  slice_head(n = 10)

# Output
write.csv(encounter, "synthetic_encounters.csv", row.names = FALSE, na = "")
  
```

Done! Enjoy your synthetic EHR data for cohort feasibility practice.

*created: 2026-06-29*

*Last updated: 2026-06-30*
