---
title: "Cohort Feasibility Assessment Using Synthetic EHR Data"
subtitle: "From three relational tables to a one-row-per-patient cohort"
description: "A reproducible SQL walkthrough for defining an index date, applying eligibility criteria, tracking cohort attrition, and summarizing a synthetic sleep disorder–Alzheimer's disease cohort."
author: "Jun Ha Chang"
date: "2026-07-20"
categories: [SQL, EHR, Clinical Data Science, Real-World Data]
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this post"
    toc-depth: 2
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    code-overflow: wrap
execute:
  warning: false
  message: false
---

::: {.callout-note appearance="simple"}
**EHR Data Workflow · Post 2**

[← Post 1: Duplicate Records in EHR Data](../before-modeling-ehr-data/index.qmd) · [View the source and synthetic data on GitHub](https://github.com/junhachang/rwe-ehr-notes/tree/main/posts/sql-participant-table)
:::

Before choosing a model—or promising one in a grant—I want to know whether the cohort actually exists in the data.

In this walkthrough, I use three small synthetic EHR tables to answer a practical feasibility question:

> Among adults with a documented sleep disorder, how many meet basic eligibility criteria for a study of incident Alzheimer's disease, and what does the final cohort look like?

The workflow produces two useful deliverables:

1. a sequential cohort attrition table; and
2. a demographic summary of the final analytic cohort.

The key design decision is simple: **reduce each source table to one row per patient before joining it to the participant master table.** This prevents a many-to-many join from quietly changing the cohort size.

## 1. Define the question before writing SQL

For this example, cohort entry is the first documented sleep disorder diagnosis. The study end date is June 30, 2026.

A patient enters the final cohort if she or he:

- has at least one sleep disorder diagnosis (`G47%`);
- is at least 50 years old at the first sleep disorder diagnosis;
- has no Alzheimer's disease diagnosis (`G30%`) on or before that index date; and
- has at least 365 days between the index date and the last observed date.

An Alzheimer's disease diagnosis after the index date and on or before the last observed date is classified as an incident event.

::: {.callout-warning appearance="simple"}
The `G47%` and `G30%` patterns are intentionally simplified for teaching. A real study needs a prespecified, clinically reviewed phenotype that defines included codes, diagnosis settings, confirmation rules, and coding systems.
:::

The 365-day follow-up rule is also a **feasibility screen**, not a ready-made survival-analysis decision. Conditioning eligibility on future observation can change the risk set. In a formal time-to-event analysis, I would replace it with a defensible observation requirement or an explicitly defined lag period.

## 2. Understand the source tables

This project uses three synthetic EHR-style tables. They are linked by `PATID`, but they do not share the same unit of analysis.

| Table | Unit of analysis | Key columns | Role in this workflow |
|---|---|---|---|
| `patients` | One row per patient | `PATID`, `BIRTH_DATE`, `DEATH_DATE`, `SEX`, `RACE` | Demographics and death date |
| `diagnoses` | Multiple rows per patient | `PATID`, `ICD10`, `ICD10_TXT`, `DX_DATE` | Sleep disorder index and Alzheimer's disease dates |
| `encounters` | Multiple rows per patient | `PATID`, `ENCOUNTER_ID`, `ENCOUNTER_DATE`, `ENCOUNTER_TYPE` | Observed EHR activity and follow-up |

```{mermaid}
erDiagram
    patients ||--o{ diagnoses : PATID
    patients ||--o{ encounters : PATID

    patients {
        string PATID PK
        date BIRTH_DATE
        date DEATH_DATE
        string SEX
        string RACE
    }

    diagnoses {
        string PATID FK
        string ICD10
        string ICD10_TXT
        date DX_DATE
    }

    encounters {
        string PATID FK
        string ENCOUNTER_ID
        date ENCOUNTER_DATE
        string ENCOUNTER_TYPE
    }
```

The files are available here: [patients](data/synthetic_patients.csv), [diagnoses](data/synthetic_diagnoses.csv), and [encounters](data/synthetic_encounters.csv). All values are synthetic.

### Load the CSV files into SQLite

The setup code creates a temporary SQLite database in memory, reads every field as text, and loads the three CSV files as relational tables. Reading ISO-formatted dates as text is intentional: SQLite's `DATE()` and `JULIANDAY()` functions can parse them consistently.

```{r}
#| label: load-synthetic-data

library(DBI)
library(RSQLite)
library(readr)

con <- dbConnect(SQLite(), ":memory:")

read_ehr_csv <- function(path) {
  read_csv(
    path,
    col_types = cols(.default = col_character()),
    na = c("", "NA")
  )
}

patients <- read_ehr_csv("data/synthetic_patients.csv")
diagnoses <- read_ehr_csv("data/synthetic_diagnoses.csv")
encounters <- read_ehr_csv("data/synthetic_encounters.csv")

dbWriteTable(con, "patients", patients, overwrite = TRUE)
dbWriteTable(con, "diagnoses", diagnoses, overwrite = TRUE)
dbWriteTable(con, "encounters", encounters, overwrite = TRUE)
```

If you are working directly in a clinical data warehouse, this loading step disappears; `con` would point to the database rather than to an in-memory SQLite instance.

## 3. Audit the source tables

I start with both record counts and unique-patient counts. A table can contain hundreds of rows without containing hundreds of patients.

```{sql, connection=con}
SELECT
    'patients' AS source_table,
    COUNT(*) AS n_rows,
    COUNT(DISTINCT PATID) AS n_unique_patients
FROM patients

UNION ALL

SELECT
    'diagnoses',
    COUNT(*),
    COUNT(DISTINCT PATID)
FROM diagnoses

UNION ALL

SELECT
    'encounters',
    COUNT(*),
    COUNT(DISTINCT PATID)
FROM encounters;
```

The synthetic files contain 100 patients, 279 diagnosis records, and 809 encounter records. The difference between rows and patients is expected because diagnoses and encounters are one-to-many tables.

I also inspect a few raw rows before transforming anything. This catches unexpected column types, inconsistent date formats, and obvious coding surprises early.

```{sql, connection=con}
SELECT *
FROM patients
LIMIT 3;
```

```{sql, connection=con}
SELECT *
FROM diagnoses
ORDER BY PATID, DX_DATE
LIMIT 5;
```

```{sql, connection=con}
SELECT *
FROM encounters
ORDER BY PATID, ENCOUNTER_DATE
LIMIT 5;
```

## 4. Reduce each event table to one row per patient

The `diagnoses` and `encounters` tables cannot be joined safely in their raw form: a patient with five diagnoses and ten encounters could produce 50 rows. Instead, I first create three patient-level views.

### First sleep disorder diagnosis

```{sql, connection=con}
#| output: false

CREATE TEMP VIEW sleep_index AS
SELECT
    PATID,
    MIN(DATE(DX_DATE)) AS INDEX_DATE
FROM diagnoses
WHERE ICD10 LIKE 'G47%'
GROUP BY PATID;
```

### First Alzheimer's disease diagnosis

```{sql, connection=con}
#| output: false

CREATE TEMP VIEW ad_index AS
SELECT
    PATID,
    MIN(DATE(DX_DATE)) AS FIRST_AD_DATE
FROM diagnoses
WHERE ICD10 LIKE 'G30%'
GROUP BY PATID;
```

### Encounter boundaries

```{sql, connection=con}
#| output: false

CREATE TEMP VIEW encounter_bounds AS
SELECT
    PATID,
    MIN(DATE(ENCOUNTER_DATE)) AS FIRST_ENCOUNTER_DATE,
    MAX(DATE(ENCOUNTER_DATE)) AS LAST_ENCOUNTER_DATE
FROM encounters
GROUP BY PATID;
```

Each view now contains no more than one row per patient. That invariant is what makes the next join predictable.

```{sql, connection=con}
SELECT 'sleep_index' AS derived_table,
       COUNT(*) AS n_rows,
       COUNT(DISTINCT PATID) AS n_unique_patients
FROM sleep_index

UNION ALL

SELECT 'ad_index', COUNT(*), COUNT(DISTINCT PATID)
FROM ad_index

UNION ALL

SELECT 'encounter_bounds', COUNT(*), COUNT(DISTINCT PATID)
FROM encounter_bounds;
```

## 5. Build the participant master table

The participant master table combines demographics, index and outcome dates, observation boundaries, and eligibility flags. Its unit of analysis is one patient—not one diagnosis or encounter.

```{mermaid}
flowchart TD
    A["Three source tables"] --> B["Patient-level dates"]
    B --> C["Participant master"]
    C --> D["Eligibility criteria"]
    D --> E["Attrition and summary"]
```

```{sql, connection=con}
#| output: false

CREATE TEMP VIEW cohort_master AS
WITH assembled AS (
    SELECT
        p.PATID,
        DATE(p.BIRTH_DATE) AS BIRTH_DATE,
        DATE(p.DEATH_DATE) AS DEATH_DATE,
        p.SEX,
        p.RACE,
        s.INDEX_DATE,
        a.FIRST_AD_DATE,
        e.FIRST_ENCOUNTER_DATE,
        e.LAST_ENCOUNTER_DATE,
        CAST(
            (JULIANDAY(s.INDEX_DATE) - JULIANDAY(p.BIRTH_DATE)) / 365.25
            AS INTEGER
        ) AS AGE_AT_INDEX,
        CASE
            WHEN p.DEATH_DATE IS NOT NULL
                 AND DATE(p.DEATH_DATE) <= DATE('2026-06-30')
                THEN DATE(p.DEATH_DATE)
            WHEN e.LAST_ENCOUNTER_DATE IS NULL
                THEN NULL
            WHEN DATE(e.LAST_ENCOUNTER_DATE) > DATE('2026-06-30')
                THEN DATE('2026-06-30')
            ELSE DATE(e.LAST_ENCOUNTER_DATE)
        END AS LAST_OBS_DATE
    FROM patients AS p
    INNER JOIN sleep_index AS s
        ON p.PATID = s.PATID
    LEFT JOIN ad_index AS a
        ON p.PATID = a.PATID
    LEFT JOIN encounter_bounds AS e
        ON p.PATID = e.PATID
),
flagged AS (
    SELECT
        *,
        CASE
            WHEN FIRST_AD_DATE IS NOT NULL
                 AND FIRST_AD_DATE <= INDEX_DATE
                THEN 1
            ELSE 0
        END AS PREVALENT_AD,
        CAST(
            JULIANDAY(LAST_OBS_DATE) - JULIANDAY(INDEX_DATE)
            AS INTEGER
        ) AS FOLLOWUP_DAYS
    FROM assembled
)
SELECT
    *,
    CASE
        WHEN FIRST_AD_DATE > INDEX_DATE
             AND FIRST_AD_DATE <= LAST_OBS_DATE
            THEN 1
        ELSE 0
    END AS AD_STATUS
FROM flagged;
```

There are several deliberate choices here:

- `INNER JOIN sleep_index` limits the master table to patients with a sleep disorder.
- `LEFT JOIN` retains those patients even if they have no Alzheimer's disease diagnosis or no encounter summary.
- death ends observation when it occurs before the study cutoff;
- otherwise, the last encounter—capped at the study end—defines the last observed date; and
- prevalent and incident Alzheimer's disease are distinguished relative to the same index date.

Here is what the patient-level output looks like:

```{sql, connection=con}
SELECT
    PATID,
    INDEX_DATE,
    AGE_AT_INDEX,
    FIRST_AD_DATE,
    LAST_OBS_DATE,
    FOLLOWUP_DAYS,
    PREVALENT_AD,
    AD_STATUS
FROM cohort_master
ORDER BY PATID
LIMIT 10;
```

For example, synthetic patient `P0001` enters the cohort on December 2, 2024, at age 72. The first Alzheimer's disease diagnosis occurs later, on January 2, 2026, and before the last observed date. The record is therefore classified as incident Alzheimer's disease (`AD_STATUS = 1`).

## 6. Apply criteria sequentially

A single final count is not enough. Sequential attrition shows which eligibility rule costs the most patients and makes the cohort definition auditable.

```{sql, connection=con}
WITH stages AS (
    SELECT
        1 AS step,
        'Patients with a sleep disorder diagnosis' AS criterion,
        COUNT(*) AS n
    FROM cohort_master

    UNION ALL

    SELECT
        2,
        'Age 50 years or older at index',
        COUNT(*)
    FROM cohort_master
    WHERE AGE_AT_INDEX >= 50

    UNION ALL

    SELECT
        3,
        'No Alzheimer''s disease on or before index',
        COUNT(*)
    FROM cohort_master
    WHERE AGE_AT_INDEX >= 50
      AND PREVALENT_AD = 0

    UNION ALL

    SELECT
        4,
        'At least 365 days of follow-up',
        COUNT(*)
    FROM cohort_master
    WHERE AGE_AT_INDEX >= 50
      AND PREVALENT_AD = 0
      AND FOLLOWUP_DAYS >= 365
)
SELECT
    step,
    criterion,
    n,
    LAG(n) OVER (ORDER BY step) - n AS excluded_at_step
FROM stages
ORDER BY step;
```

The sample narrows from 55 patients with a sleep disorder diagnosis to 27 patients in the final cohort. The table also shows exactly where the other 28 patients were excluded.

Now I save the final eligibility logic as another patient-level view.

```{sql, connection=con}
#| output: false

CREATE TEMP VIEW sleep_ad_cohort AS
SELECT *
FROM cohort_master
WHERE AGE_AT_INDEX >= 50
  AND PREVALENT_AD = 0
  AND FOLLOWUP_DAYS >= 365;
```

## 7. Summarize the feasible cohort

For feasibility, I want a compact answer: total patients, outcome events, age, sex, race, and follow-up. I do not need a publication-ready Table 1 yet.

```{sql, connection=con}
SELECT
    CASE AD_STATUS
        WHEN 1 THEN 'Incident AD'
        ELSE 'No incident AD'
    END AS outcome_group,
    COUNT(*) AS n,
    ROUND(AVG(AGE_AT_INDEX), 1) AS mean_age,
    SUM(CASE WHEN SEX = 'Female' THEN 1 ELSE 0 END) AS female_n,
    ROUND(
        100.0 * SUM(CASE WHEN SEX = 'Female' THEN 1 ELSE 0 END) / COUNT(*),
        1
    ) AS female_pct,
    ROUND(AVG(FOLLOWUP_DAYS) / 365.25, 1) AS mean_followup_years
FROM sleep_ad_cohort
GROUP BY AD_STATUS
ORDER BY AD_STATUS;
```

```{sql, connection=con}
SELECT
    CASE AD_STATUS
        WHEN 1 THEN 'Incident AD'
        ELSE 'No incident AD'
    END AS outcome_group,
    RACE,
    COUNT(*) AS n,
    ROUND(
        100.0 * COUNT(*)
        / SUM(COUNT(*)) OVER (PARTITION BY AD_STATUS),
        1
    ) AS pct_within_group
FROM sleep_ad_cohort
GROUP BY AD_STATUS, RACE
ORDER BY AD_STATUS, RACE;
```

The final synthetic cohort contains 27 patients and only 2 incident Alzheimer's disease events. That is **not enough for a meaningful adjusted time-to-event model**. This is not a failed workflow; it is the point of a feasibility assessment. The screen tells us to reconsider the data source, study window, outcome definition, or analytic ambition before investing in a model that the data cannot support.

## 8. Run patient-level quality checks

Before exporting the cohort, I verify that the final table still has one row per patient and that core dates are plausible.

```{sql, connection=con}
SELECT
    COUNT(*) AS n_rows,
    COUNT(DISTINCT PATID) AS n_unique_patients,
    COUNT(*) - COUNT(DISTINCT PATID) AS duplicate_patient_rows,
    SUM(CASE WHEN INDEX_DATE IS NULL THEN 1 ELSE 0 END) AS missing_index_dates,
    SUM(CASE WHEN AGE_AT_INDEX < 0 THEN 1 ELSE 0 END) AS impossible_ages,
    SUM(CASE WHEN FOLLOWUP_DAYS < 0 THEN 1 ELSE 0 END) AS negative_followup
FROM sleep_ad_cohort;
```

For this example, the final table has 27 rows and 27 unique patients, with no missing index dates, impossible ages, or negative follow-up values.

These checks are intentionally basic. In a real EHR study, I would also inspect diagnosis provenance, repeated code requirements, encounter gaps, death-date completeness, out-of-network care, and whether the chosen observation rule has the same meaning across sites.

## What to take away

The most important output is not the final SQL view. It is the chain of decisions that makes the view interpretable:

1. define the research question and time origin;
2. identify the unit of analysis in each source table;
3. reduce event tables to one row per patient before joining;
4. apply eligibility rules sequentially;
5. count outcome events—not only patients; and
6. stop when the data do not support the planned analysis.

That last step is easy to skip. A cohort can be large enough to look impressive and still contain too few events for the proposed model.

### SQL dialect note

This post uses SQLite. Other database systems use different date functions—for example, `DATEDIFF()` rather than `JULIANDAY()`—but the cohort logic is portable. Translate the date syntax, not the study design.

::: {.callout-important appearance="simple"}
**Data disclosure:** All examples are synthetic and were created for educational purposes. They contain no real patient records, identifiers, institutional data, or results from a specific health system.
:::

```{r}
#| label: disconnect
#| include: false

dbDisconnect(con)
```
